---
title: "External_Tick"
author: "Carrie Vanty"
date: "2025-04-02"
output: html_document
---

First, we must load our data
```{r, echo=FALSE}
dat <- read.csv("monthly_tick_temp.csv")
library(rjags)
```
The data is the tick and temperature data used in the other models. We only used data if we had the month/year available for both tick count and average monthly temperature. If there were two tick counts for a given month/year, we averaged them together so that we had an average for the month.

formatting data:
```{r}
t = dat[1:58,1] #time
tick = round(dat[1:58,3]) #monthly average ticks
temp = dat[1:58,2] #monthly average temperature
data <- list(tick = tick, temp = temp, n = length(t))

data$temp <- round(data$temp) 
```


```{r}
plot(temp,tick)
```
```{r}
mod <- lm(tick~temp)
summary(mod)
```
#Process Model

We chose a linear model as our process model, which produces the mean tick count for a given temperature. Epsilon is our process error that is distributed normally.

$$\mu_{tick-count}[i] = \beta_1 + \beta_2temp[i] + \epsilon[i]$$
$$\epsilon \sim N(0, \tau)$$

#Data Model

We chose to use the Poisson distribution for our data model, since it is discrete and always positive. 

$$y[i] \sim Pois(\mu_{tick-count}[i])$$

#Choosen Priors

We are using uninformed priors for our process model coefficients.

$$\beta_1 \sim N(0,1/1,000)$$
$$\beta_2 \sim N(0,1/1,000)$$
For tau (the precision of our process error), we are using the gamma distribution, which is the general choice for generating precision.

$$\tau \sim \Gamma(0.1,0.1)$$


Specify Model
```{r}
tick_external <- "
model{

  b1 ~ dnorm(b0,1/1000)  
  b2 ~ dnorm(b0,1/1000)    ## multivariate Normal prior on vector of regression params
  
  tau ~ dgamma(0.1,0.1)  ## precision for process error
  

  for(i in 1:n){
      e[i] ~ dnorm(0,tau)             ## process error
      mu[i] <- b1 + b2*temp[i] + e[i] ## process model mu = mean monthly tick
      #y[i]  ~ dnorm(mu[i],tau)  
      tick[i]  ~ dpois(max(0.001, mu[i]))              ## data model
  }
}
"
```

Specify priors
```{r}
data$b0 <- 0     ## regression b means
#data$y <- tick
```

Set up initial conditions
```{r}
inits <- list()

nchain = 3
inits <- list()
for(i in 1:nchain){
 inits[[i]] <- list(b1 = rnorm(1,0,5), b2 = rnorm(1,0,5))
}
```


Call jags model
```{r}
j.model <- jags.model(file = textConnection(tick_external),
                      data = data,
                      inits = inits,
                      n.chains = 3)
```

```{r}
jags.out <- coda.samples(model = j.model,
                         variable.names = c("b1","b2","tau"),
                         n.iter = 75000)
```

Plot
```{r}
plot(jags.out)
```
The trace plot for tau appears to have converged since the chains are overlapping. The trace plots of b1 and b2 do have overlap, but not as strong as tau, indicating that it may not have converged yet. 

Summary statistics
```{r}
summary(jags.out)
```
b2 is around three, which is the value R's linear model found. b1 is around -12 and should theoretically be around -7

Correlation of parameters
```{r}
out <- as.matrix(jags.out)
pairs(out)
cor(out) 
```
b1 and b2 appear to be correlated now, which is what we want from a linear model. We also see no correlation with b1,b2 and tau, which is what we expect.

```{r}
effectiveSize(jags.out)
```
Gelman Plots
```{r}
gelman.plot(jags.out)
gelman.diag(jags.out)
```
```{r}
burnin = 15000                                   ## determine convergence
jags.burn <- window(jags.out, start = burnin)  ## remove burn-in
plot(jags.burn)       ## check diagnostics post burn-in
```
Our chains are not overlapping as much as they should if they are converging. This is likely due to our low precision.

We also wanted to consider the summary, plots and convergence of the predicted mean number of ticks.
```{r}
jags.out2 <- coda.samples(model = j.model,
                         variable.names = c("mu[10]","mu[5]"),
                         n.iter = 60000)

summary(jags.out2)
```


Plot
```{r}
plot(jags.out2)
```
We do have overlapping chains in our trace plots for mu indicating that the values are converging. The values are also appropriate for the number of ticks at our site.

```{r}
gelman.plot(jags.out2)
gelman.diag(jags.out2)
```
 We have convergence for mu values tested, thus we do not need burn in.



