---
title: "External_Tick"
author: "Carrie Vanty"
date: "2025-04-02"
output: html_document
---

First, we must load our data
```{r, echo=FALSE}
dat <- read.csv("monthly_tick_temp.csv")
library(rjags)
```
The data is the tick and temperature data used in the other models. We only used data if we had the month/year available for both tick count and average monthly temperature. If there were two tick counts for a given month/year, we averaged them together so that we had an average for the month.

formatting data:
```{r}
t = dat[1:58,1] #time
tick = dat[1:58,3] #monthly average ticks
temp = dat[1:58,2] #monthly average temperature
data <- list(temp = temp, n = length(t))

data$temp <- round(data$temp) 
```


```{r}
plot(temp,tick)
```

#Process Model

We chose a linear model as our process model, which produces the mean tick count for a given temperature. Epsilon is our process error that is distributed normally.

$$\mu_{tick-count}[i] = \beta_0 + \beta_1temp[i] + \epsilon[i]$$
$$\epsilon \sim N(0, \tau)$$

#Data Model

We chose to use the Poisson distribution for our data model, since it is discrete and always positive. 

$$y[i] \sim Pois(\mu_{tick-count}[i])$$

#Choosen Priors

We are using uninformed priors for our process model coefficients.

$$\beta_1 \sim N(0,1/1,000)$$
$$\beta_2 \sim N(0,1/1,000)$$
For tau (the precision of our process error), we are using the gamma distribution, which is the general choice for generating precision.

$$\tau \sim \Gamma(0.1,0.1)$$


Specify Model
```{r}
tick_external <- "
model{

  b1 ~ dnorm(b0,1/1000)  
  b2 ~ dnorm(b0,1/1000)    ## multivariate Normal prior on vector of regression params
  
  tau ~ dgamma(0.1,0.1)  ## precision for process error
  

  for(i in 1:n){
      e[i] ~ dnorm(0,tau)             ## process error
      mu[i] <- b1 + b2*temp[i] + e[i]  ## process model mu = mean monthly tick
      y[i]  ~ dpois(max(0.001, mu[i]))              ## data model
  }
}
"
```

Specify priors
```{r}
data$b0 <- 0     ## regression b means
```

Letting JAGS decide on initial conditions, so that they are uninformed

Call jags model
```{r}
j.model <- jags.model(file = textConnection(tick_external),
                      data = data,
                      n.chains = 3)
```

```{r}
jags.out <- coda.samples(model = j.model,
                         variable.names = c("b1","b2","tau"),
                         n.iter = 10000)
```

Plot
```{r}
plot(jags.out)
```
The trace plots for b1 and b2 appear to have converged since the chains are overlapping. The trace of tau does have overlap, but not as strong as b1 and b2, indicating that it may not have converged yet. Additionally, tau is spending a lot of time around zero, which is likely due to our large precision.

Summary statistics
```{r}
summary(jags.out)
```


Correlation of parameters
```{r}
out <- as.matrix(jags.out)
pairs(out)
cor(out) 
```
b1 and b2 appear to be uncorrelated. The correlation values associated with tau are small, but the graphs appear different than the typical correlation plots due to tau spending a lot of time near zero.

```{r}
effectiveSize(jags.out)
```
Gelman Plots
```{r}
gelman.plot(jags.out)
gelman.diag(jags.out)
```
We have convergence for b1,b2, and tau since all values from the Gelman diagnostics are equal to one.

We also wanted to consider the summary, plots and convergence of the predicted mean number of ticks.

```{r}
jags.out2 <- coda.samples(model = j.model2,
                         variable.names = c("mu[10]","mu[5]"),
                         n.iter = 60000)

gelman.diag(jags.out2)
```


Plot
```{r}
plot(jags.out2)
```
We do not have overlapping chains in our trace plots for mu indicating that the values are not converging.

```{r}
gelman.plot(jags.out2)
gelman.diag(jags.out2)
```
 Even with an extremely high iteration rate (went up to 20,000,000) and more chains, we did not get convergence for the mu values. This is likely because we have high variance for b1 and b2, so the chains that the BGR statistic are measuring may not converge to the same value. For our next model, we will try and find information to establish an informed prior, so the variance is smaller.

```{r}
burnin = 10000                                   ## determine convergence
jags.burn <- window(jags.out2, start = burnin)  ## remove burn-in
plot(jags.burn)       ## check diagnostics post burn-in
```

